---
title: "Data Visualization"
author: "Geetha"
date: "23/08/2020"
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
                     
library(ggplot2)
library(dplyr)
library(countrycode)
library(gsheet)
library(tidyr)
library(readxl)
library(stringr)
library(GGally)

```

```{r dataset, include=FALSE, warning=FALSE}
#GLOBAL DATASETS
#####################
#Our World in Data
owid <- read.csv("https://covid.ourworldindata.org/data/owid-covid-data.csv")

owid <- (owid %>% 
        mutate(location = as.factor(location), continent = as.factor(continent), iso_code = as.factor(iso_code), date = (as.Date(date, format= "%Y-%m-%d")),tests_units = as.factor(tests_units)) %>% 
        mutate(cumulative.tpr = (total_cases/total_tests)*100, daily.tpr = (new_cases/new_tests)*100) %>% 
        filter(!is.na(daily.tpr)) %>% 
        select (!(total_cases:stringency_index)) %>%  
        rename(country = location, ISO = iso_code))
        

#DELVE (Data Evaluation and Learning for Viral Epidemics) Dataset 
delve = read.csv(url("https://raw.githubusercontent.com/rs-delve/covid19_datasets/master/dataset/combined_dataset_latest.csv"))
delve <- (delve %>%
            mutate(DATE = as.Date(DATE, format= "%Y-%m-%d"), country_name = as.factor(country_name), ISO = as.factor(ISO)) %>%
            mutate(across(c(starts_with("npi") & !(npi_fiscal_measures|npi_international_support|npi_healthcare_investment|npi_vaccine_investment|npi_stringency_index)), as.factor)) %>% 
            rename (date = DATE, country = country_name))
          


#Country Effective Reproductive Number Estimates from University of Oxford, Australian National University, and Harvard from http://epidemicforecasting.org/
Rt <- read.csv("https://storage.googleapis.com/static-covid/static/v4/main/r_estimates.csv")
Rtcountry <- (Rt %>%
                      mutate(Date = as.Date(Date, format= "%Y-%m-%d"), X = NULL) %>%
                      filter(!grepl ("^US-", Code), !grepl ("^AU-", Code), !grepl ("^CN-", Code)) %>% 
                      mutate(country = countrycode(Code, "iso2c","country.name"), country =as.factor(country)) %>% #converting country codes to country names. States not converted 
                      rename(date=Date))

#Open Data Barometer (Openness of data measurement)
odb <- read.csv("https://opendatabarometer.org/assets/data/ODB-2014-Rankings.csv")
odb <- odb %>% select (ISO3:ODB.Scaled) %>% rename(country = Country, ISO = ISO3)

#global health security index (from https://www.ghsindex.org/wp-content/uploads/2019/10/Global-Health-Security-Index-2019-Final-October-2019.zip) *will update with better link once .xlsm data import figured out
ghsrange <- "N4:OP267"
ghslast <- 197
# importing the iScores sheet of the xlsm file,
GHSI <- read_excel("ghsindex.xlsm",sheet = "iScores", range = ghsrange,col_names = TRUE) 
GHSI <- t(GHSI)
#making the first row the column names
ghscoltitle <- GHSI[1,] 
colnames(GHSI, do.NULL = TRUE, prefix = "col" )
colnames(GHSI) <- ghscoltitle
#turn excel matrix into data frame
GHSI <- as.data.frame(GHSI)
names(GHSI) <- (names(GHSI) %>% str_replace_all("\\s","_"))
GHSI <- (GHSI %>%
                 rename(country = Indicators) %>% 
                 rename( '1)Prevent' = `1)_PREVENTION_OF_THE_EMERGENCE_OR_RELEASE_OF_PATHOGENS`, '2)Detect' = `2)_EARLY_DETECTION_&_REPORTING_EPIDEMICS_OF_POTENTIAL_INTERNATIONAL_CONCERN`, '3)Response' = `3)_RAPID_RESPONSE_TO_AND_MITIGATION_OF_THE_SPREAD_OF_AN_EPIDEMIC`,'4)Health' = `4)_SUFFICIENT_&_ROBUST_HEALTH_SECTOR_TO_TREAT_THE_SICK_&_PROTECT_HEALTH_WORKERS`,'5)Norms' = `5)_COMMITMENTS_TO_IMPROVING_NATIONAL_CAPACITY,_FINANCING_AND_ADHERENCE_TO_NORMS`, '6)Risk' = `6)_OVERALL_RISK_ENVIRONMENT_AND_COUNTRY_VULNERABILITY_TO_BIOLOGICAL_THREATS`) %>% 
                 slice(2:ghslast) %>% 
                 mutate_at(vars(-country), as.numeric))

# Putting together global Dataset
global <- full_join(owid, delve, by = c("country", "date"))
global <- full_join(global,Rtcountry, by = c("country", "date")) 
global <- full_join (global, odb, by = "country", "ISO")
global <- full_join(global, GHSI, by = "country")

global <- (global %>% 
                   mutate(ISO = as.factor(ISO), country = as.factor(country)))
```

```{r reduced, echo=FALSE}
#Filtered Dataset
OECD <- c("United States", "Mexico", "Japan", "Turkey", "Germany", "United Kingdom", "France", "Italy", "South Korea","Spain", "Poland", "Canada", "Australia", "Chile", "Netherlands", "Belgium", "Czech Republic", "Greece", "Portugal", "Sweden", "Hungary", "Austria", "Israel", "Switzerland", "Denmark", "Finland", "Slovakia", "Norway", "Ireland"," New Zealand", "Lithuania", "Slovenia", "Latvia","Estonia", "Luxembourg", "Iceland")
reduced <- (global %>%  filter(country == OECD))
```

## Overview of Global Performance 
South Korea, Vietnam, Germany and New Zealand have been highlighted by [Exemplars in Global Health](https://www.exemplars.health/emerging-topics/epidemic-preparedness-and-response/covid-19) as countries that are doing comparatively well in managing their outbreak. Mexico, India, Indonesia, and Argentina are examples of countries that have not yet "bent the curve", many with quite high test positivity rates. 

#### Metrics

* Test positivity rate: daily TPR, cumulative TPR
* Number of Deaths (daily, cumulative, per million, per thousand, smoothed over 7 days)
* Number of cases (daily, cumulative per million, per thousand, smoothed over 7 days)
* Number of tests (raw,per million, per thousand, smoothed over 7 days) 

Potential Independent variables: 

* Non-pharmaceutical interventions: closings, masks, lockdown, debt relief, etc. 
* GHS index: overall score, category scores, sub-indicator scores
* Country characteristics: population, pop density, median age, % pop > 65 y/0, % pop > 70 y/o, GDP
* Disease burden: cardiovascular disease, diabetes, smoking,  life expectancy
* Weather: temp, wind, humidity, precipitation


Potential Options for Inclusion/Exclusion Criteria: 

* remove countries with > 100 number of confirmed cases (some papers have used 1000)
* remove countries with  population >1,000,000 
* only use OECD countries (this is what is currently used for the graphs below)
* only WHO member states 
* completeness of dataset (must have test data for X% of sample period)


Below, independent variables that were not time dependent (health indices) 

# Metrics Over Time

### Testing

#### Number of Tests
```{r testing}
newtests <- ggplot(reduced, aes(x = date, y = tests_new_per_thousand, group=country)) + 
  geom_line(aes(color=country)) + xlab("Date") + 
  ylab("New Tests Per Thousand") +
  ggtitle("New Tests (per thousand people) over Time") + 
  labs(color="Country") +
  theme(plot.title = element_text(hjust = 0.5))
print(newtests)

```

#### Test Positivity Rates Over Time
```{r ctpr, warning=FALSE, echo=FALSE}
# Plot 1: Line graph to visualize/compare daily TPR over time
ctpr <- ggplot(reduced, aes(x = date, y = cumulative.tpr, group=country)) + 
  geom_line(aes(color=country)) + xlab("Date") + 
  ylab("Cumulative Test Positivity Rate (%)") +
  ggtitle("Cumulative TPR Over Time") + 
  labs(color="Country") +
  theme(plot.title = element_text(hjust = 0.5))
print(ctpr)
```

```{r dtpr}
dtpr <- ggplot(reduced, aes(x = date, y = daily.tpr, group=country)) + 
  geom_line(aes(color=country)) + xlab("Date") + 
  ylab("Daily Test Positivity Rate (%)") +
  ggtitle("Daily TPR Over Time") + 
  labs(color="Country") +
  theme(plot.title = element_text(hjust = 0.5))
print(dtpr)
```


#### Number of Cases
```{r cases}
newcases <- ggplot(reduced, aes(x = date, y = cases_new_per_million, group=country)) + 
  geom_line(aes(color=country)) + xlab("Date") + 
  ylab("New Cases Per Million") +
  ggtitle("Number of New Cases per Million People") + 
  labs(color="Country") +
  theme(plot.title = element_text(hjust = 0.5))
print(newcases)
```

#### Number of Deaths
```{r deaths}
newdeaths <- ggplot(reduced, aes(x = date, y = deaths_new_per_million, group=country)) + 
  geom_line(aes(color=country)) + xlab("Date") + 
  ylab("New Deaths Per Million") +
  ggtitle("Number of New Deaths per Million People") + 
  labs(color="Country") +
  theme(plot.title = element_text(hjust = 0.5))
print(newdeaths)
```


## Non Pharmaceutical Interventions
##### Overal Government Stringency Score
```{r NPI, echo=FALSE}
npi <- ggplot(reduced, aes(x= npi_stringency_index, y=cases_new_per_million, 
                      color=country))+ 
  geom_boxplot()+
  labs(col="Country")+
  xlab("NPI Stringency Index") + ylab("New Cases/Million People") + 
  ggtitle("New Cases and NPI Stringency Index") +
  theme(plot.title = element_text(hjust = 0.5))
print(npi)
```

### Interventions within Stringency Index
```{r NPI interventions}
npi_int <- ggparcoord(reduced, columns = c(22:29, 35, 36, 40),
                  groupColumn = "cases_new_per_million", scale="uniminmax", alphaLines=0.1) +
  xlab("NPI Category") + ylab("Value (scaled between 0 and 1") + 
  ggtitle("Effect of NPIs on New Cases/Million") +
  theme_minimal()+ theme(plot.title = element_text(hjust = 0.5)) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(npi_int)
```

### Mask Policy 
Intensity of mask policy within region.

1 - General recommendation

2 - Mandatory in specific regions of country or specific places (public transport, inside shops)

3 - Mandatory everywhere or universal usage

```{r masks, echo=FALSE}
mask <- ggplot(reduced, aes(x=npi_masks, y=cumulative.tpr))+ 
  geom_boxplot()+
  xlab("Mask Policy") + ylab("Cumulative Test Positivity Rate") + 
  ggtitle("Mask Policy and Cumulative Test Positivity Rate") +
  theme(plot.title = element_text(hjust = 0.5))
print(mask)
```


## Disease Burden
Evaluating the effects of prevalence of diabetes (% of population), smoking (% of group), and cardiovascular deaths (per 1000 people) on the cumulative test positivity rate. 
```{r disease}
# Disease burden factors and cTPR
db <- ggparcoord(reduced, columns = 12:15,
                  groupColumn = "cumulative.tpr", scale="uniminmax", alphaLines=0.1) +
  xlab("Disease burden") + ylab("Value") + ggtitle("Disease burden and cumulative TPR") +
  theme_minimal()+ theme(plot.title = element_text(hjust = 0.5)) + theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
print(db)
```



## Mobility Measures
Mobility Trends in % change relative to baseline, from Google Mobility Reports. 
```{r mobility, warning=FALSE, echo=FALSE}
pcp <- ggparcoord(reduced, columns = 64:72,
                  groupColumn = "cases_total_per_million", scale="uniminmax", alphaLines=0.1) +
  xlab("mobility factors") + ylab("Value") + 
  ggtitle("Mobility factors and total cases per million people") +
  theme(plot.title = element_text(hjust = 0.5)) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(pcp)
```


## Global Health Security Index
```{r ghs index}

# Plot 1: Line graph to visualize/compare daily TPR over time
ghs <- ggparcoord(reduced, columns = c(19, 20, 45, 47, 51),
                  groupColumn = "OVERALL_SCORE", scale="uniminmax", alphaLines=0.1) +
  xlab("COVID-19 Data") + ylab("Scaled Values") + ggtitle("GHS Index vs outbreak/testing indicators") +
  labs(col = c("1","2")) +
  theme_minimal()+ theme(plot.title = element_text(hjust = 0.5)) + theme(axis.text.x = element_text(angle = 10, hjust = 0.5)) 
print(ghs)

```

### GHSI Categories
```{r ghs categories}

# Plot 1: Line graph to visualize/compare daily TPR over time
ghs <- ggparcoord(reduced, columns = c(97, 154, 191, 234, 273, 312),
                  groupColumn = "cumulative.tpr", scale="uniminmax", alphaLines=0.1) +
  xlab("GHS Indicator") + ylab("Scaled Indicator Value") + ggtitle("Cumulative TPR versus GHS") +
  labs(col = c("1","2")) +
  theme_minimal()+ theme(plot.title = element_text(hjust = 0.5)) 
print(ghs)
```




### Healthcare Access
Assessing relationship between healthcare access (measured by ghs indices) and deaths
```{r}
ghs <- ggparcoord(reduced, columns = c(237, 241, 242, 250, 252),
                  groupColumn = "deaths_total_per_million", scale="uniminmax", alphaLines=0.1) +
  xlab("Healthcare Access Indicators") + ylab("Scaled Values") + ggtitle("Deaths per million vs healthcare access measures") +
  labs(col = c("1","2")) +
  theme_minimal()+ theme(plot.title = element_text(hjust = 0.5)) + theme(axis.text.x = element_text(angle = 15, hjust = 0.5)) + scale_x_discrete(labels = function(x) str_wrap(x, width = 5 ))
print(ghs)
```
 


## Weather Variables
#### Humidity
```{r}
# Humidity and total cases (upward trend)
hum <- ggplot(reduced, aes(x=weather_humidity_mean,y=cases_total_per_million))+ 
  geom_point(alpha=0.5) +
  geom_smooth(method = "lm") +
  xlab("Mean humidity") + ylab("Total cases/million people") + 
  ggtitle("Mean humidity and total cases/million people") +
  theme(plot.title = element_text(hjust = 0.5))
print(hum)
```

#### Temperature Speed

```{r}
temp <- ggplot(reduced, aes(x=weather_temperature_mean,y= cases_new_per_million))+ 
  geom_point(alpha=0.5) +
  geom_smooth(method = "lm") +
  xlab("Mean temperature") + ylab("New cases/million people") + 
  ggtitle("Mean temperature and new cases/million people") +
  theme(plot.title = element_text(hjust = 0.5))
print(temp)
```

#### Wind Speed

```{r}
wind <- ggplot(reduced, aes(x=weather_wind_speed_mean, y= cases_new_per_million))+ 
  geom_point(alpha=0.5) +
  geom_smooth(method = "lm") +
  xlab("Mean wind speed") + ylab("New cases/million people") + 
  ggtitle("Mean wind speed and new cases/million people") +
  theme(plot.title = element_text(hjust = 0.5))
print(wind)
```

