---
title: "Data Visualization"
author: "Geetha"
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, fig.width = 12,fig.height = 6)
                     
library(ggplot2)
library(dplyr)
library(countrycode)
library(gsheet)
library(tidyr)
library(readxl)
library(stringr)
library(GGally)
library(directlabels)
library(grid)

```

```{r dataset, include=FALSE, warning=FALSE}
#GLOBAL DATASETS
#####################
#Our World in Data
owid <- read.csv("https://github.com/owid/covid-19-data/raw/master/public/data/owid-covid-data.csv")

owid <- (owid %>% 
                 mutate(location = as.factor(location), continent = as.factor(continent), iso_code = as.factor(iso_code), date = (as.Date(date, format= "%Y-%m-%d")),tests_units = as.factor(tests_units)) %>% 
                 mutate(cumulative.tpr = (total_cases/total_tests)*100, daily.tpr = (new_cases/new_tests)*100) %>% 
                 rename(country = location, ISO = iso_code, daily.tpr.7dayrolling = positive_rate))
        
summary(owid)


#DELVE (Data Evaluation and Learning for Viral Epidemics) Dataset 
delve = read.csv("https://raw.githubusercontent.com/rs-delve/covid19_datasets/master/dataset/combined_dataset_latest.csv")
delve <- (delve %>%
                  mutate(DATE = as.Date(DATE, format= "%Y-%m-%d"), country_name = as.factor(country_name), ISO = as.factor(ISO)) %>%
                  mutate(across(c(starts_with("npi") & !(npi_fiscal_measures|npi_international_support|npi_healthcare_investment|npi_vaccine_investment|npi_stringency_index)), as.factor)) %>% 
                  select (!(cases_total:stats_gdp_per_capita)) %>%
                  rename (date = DATE, country = country_name))
          
summary(delve)



#Country Effective Reproductive Number Estimates from University of Oxford, Australian National University, and Harvard from http://epidemicforecasting.org/
Rt <- read.csv("https://storage.googleapis.com/static-covid/static/v4/main/r_estimates.csv")
Rtcountry <- (Rt %>%
                      mutate(Date = as.Date(Date, format= "%Y-%m-%d"), X = NULL) %>%
                      filter(!grepl ("^US-", Code), !grepl ("^AU-", Code), !grepl ("^CN-", Code)) %>% 
                      mutate(country = countrycode(Code, "iso2c","country.name"), country =as.factor(country)) %>% #converting country codes to country names. States not converted 
                      rename(date=Date))
summary(Rtcountry)

#Open Data Barometer (Openness of data measurement)
odb <- read.csv("https://opendatabarometer.org/assets/data/ODB-2014-Rankings.csv")
odb <- (odb %>% 
                select (ISO3:ODB.Scaled) %>%
                rename(country = Country, ISO = ISO3, odb_readiness = Readiness, odb_implementation = Implementation, odb_impact = Impact))
summary(odb)

#global health security index (from https://www.ghsindex.org/wp-content/uploads/2019/10/Global-Health-Security-Index-2019-Final-October-2019.zip) *will update with better link once .xlsm data import figured out
ghsrange <- "N4:OP267"
ghslast <- 197
# importing the iScores sheet of the xlsm file,
GHSI <- read_excel("ghsindex.xlsm",sheet = "iScores", range = ghsrange,col_names = TRUE) 
GHSI <- t(GHSI)
#making the first row the column names
ghscoltitle <- GHSI[1,] 
colnames(GHSI, do.NULL = TRUE, prefix = "col" )
colnames(GHSI) <- ghscoltitle
#turn excel matrix into data frame
GHSI <- as.data.frame(GHSI)
names(GHSI) <- (names(GHSI) %>% str_replace_all("\\s","_"))
GHSI <- (GHSI %>%
                 rename(country = Indicators) %>% 
                 rename( '1)prevent' = `1)_PREVENTION_OF_THE_EMERGENCE_OR_RELEASE_OF_PATHOGENS`, '2)Detect' = `2)_EARLY_DETECTION_&_REPORTING_EPIDEMICS_OF_POTENTIAL_INTERNATIONAL_CONCERN`, '3)Response' = `3)_RAPID_RESPONSE_TO_AND_MITIGATION_OF_THE_SPREAD_OF_AN_EPIDEMIC`,'4)Health' = `4)_SUFFICIENT_&_ROBUST_HEALTH_SECTOR_TO_TREAT_THE_SICK_&_PROTECT_HEALTH_WORKERS`,'5)Norms' = `5)_COMMITMENTS_TO_IMPROVING_NATIONAL_CAPACITY,_FINANCING_AND_ADHERENCE_TO_NORMS`, '6)Risk' = `6)_OVERALL_RISK_ENVIRONMENT_AND_COUNTRY_VULNERABILITY_TO_BIOLOGICAL_THREATS`) %>% 
                 slice(2:ghslast) %>% 
                 mutate_at(vars(-country), as.numeric))

# Putting together global Dataset
global <- full_join(owid, delve, by = c("country", "date", "ISO"))
global <- full_join(global,Rtcountry, by = c("country", "date")) 
global <- full_join (global, odb, by = c("country", "ISO"))
global <- full_join(global, GHSI, by = c("country"))


global <- (global %>% 
                   mutate(ISO = as.factor(ISO), country = as.factor(country), stringency_index = NULL))
```

```{r reduced, echo=FALSE}
#Filtered Dataset
OECD <- c("United States", "Mexico", "Japan", "Turkey", "Germany", "United Kingdom", "France", "Italy", "South Korea","Spain", "Poland", "Canada", "Australia", "Chile", "Netherlands", "Belgium", "Czech Republic", "Greece", "Portugal", "Sweden", "Hungary", "Austria", "Israel", "Switzerland", "Denmark", "Finland", "Slovakia", "Norway", "Ireland"," New Zealand", "Lithuania", "Slovenia", "Latvia","Estonia", "Luxembourg", "Iceland")
reduced <- (global %>%  filter(country == OECD))
```

## Overview of Global Performance 
South Korea, Vietnam, Germany and New Zealand have been highlighted by [Exemplars in Global Health](https://www.exemplars.health/emerging-topics/epidemic-preparedness-and-response/covid-19) as countries that are doing comparatively well in managing their outbreak. Mexico, India, Indonesia, and Argentina are examples of countries that have not yet "bent the curve", many with quite high test positivity rates. The graphs below only include the 36 [Organisation for Economic Co-operation and Development (OECD) Member States](https://www.oecd.org/about/members-and-partners/).

#### Metrics

* Test positivity rate: daily TPR, cumulative TPR
* Number of Deaths (daily, cumulative, per million, per thousand, smoothed over 7 days)
* Number of cases (daily, cumulative, per million, per thousand, smoothed over 7 days)
* Number of tests (daily, cumulative, per million, per thousand, smoothed over 7 days) 

Potential Independent variables: 

* Non-pharmaceutical interventions: closings, masks, lockdown, debt relief, etc. 
* GHS index: overall score, category scores, sub-indicator scores
* Country characteristics: population, pop density, median age, % pop > 65 y/0, % pop > 70 y/o, GDP
* Disease burden: cardiovascular disease, diabetes, smoking,  life expectancy
* Weather: temp, wind, humidity, precipitation


Potential Options for Inclusion/Exclusion Criteria: 

* remove countries with > 100 number of confirmed cases (some papers have used 1000)
* remove countries with  population >1,000,000 
* only use OECD countries
* only WHO member states 
* completeness of dataset (must have test data for X% of sample period)


Below, independent variables that were not time dependent (i.e. healthcare access indices) were evaluated with cumulative tests/cases/deaths, while time dependent variables (i.e. changes in non pharmaceutical interventions) were evaluated with daily rates and cumulative rates. Line plots and parallel coordinate plots were primarily used to evaluate any correlations.

# Metrics Over Time

## Testing
```{r testing}
newtests <- ggplot(reduced, aes(x = date, y = new_tests_per_thousand, group=country)) + 
  geom_line(aes(color=country)) + xlab("Date") + 
  ylab("New Tests Per Thousand") +
  ggtitle("New Tests (per thousand people) over Time") + 
  labs(color="Country") + geom_dl(aes(label = country), method = list(dl.combine("angled.endpoints")), cex = 0.8) +
  theme(plot.title = element_text(hjust = 0.5))
print(newtests)

```

```{r ctpr, warning=FALSE, echo=TRUE}
# Plot 1: Line graph to visualize/compare daily TPR over time
ctpr <- ggplot(reduced, aes(x = date, y = cumulative.tpr, group=country)) + 
  geom_line(aes(color=country)) + xlab("Date") + 
  ylab("Cumulative Test Positivity Rate (%)") +
  ggtitle("Cumulative TPR Over Time") + 
  labs(color="Country") +
  theme(plot.title = element_text(hjust = 0.5)) + 
  geom_dl(aes(label = country), method = list(dl.combine("last.polygons")))
print(ctpr)
```

```{r dtpr, , echo=TRUE}
dtpr <- ggplot(reduced, aes(x = date, y = daily.tpr.7dayrolling, group=country)) + 
  geom_line(aes(color=country)) + xlab("Date") + 
  ylab("Daily Test Positivity Rate (Rolling 7-day average %)") +
  ggtitle("Daily Test Positivity Rate Over Time") + 
  labs(color="Country") +
  theme(plot.title = element_text(hjust = 0.5))
print(dtpr)
```


## Number of Cases
```{r cases, echo=TRUE}
newcases <- ggplot(reduced, aes(x = date, y = new_cases_per_million, group=country)) + 
  geom_line(aes(color=country)) + xlab("Date") + 
  ylab("New Cases Per Million") +
  ggtitle("Number of New Cases per Million People") + 
  labs(color="Country") +
  theme(plot.title = element_text(hjust = 0.5)) +
  
print(newcases)
```

## Number of Deaths

```{r deaths, echo=TRUE}
newdeaths <- ggplot(reduced, aes(x = date, y = new_deaths_per_million, group=country)) + 
  geom_line(aes(color=country)) + xlab("Date") + 
  ylab("New Deaths Per Million") +
  ggtitle("Number of New Deaths per Million People") + 
  labs(color="Country") +
  theme(plot.title = element_text(hjust = 0.5))
print(newdeaths)
```

## Country Characteristics

```{r population density, echo=TRUE}
pop.den <- ggplot(reduced, aes(x=population_density,y=total_cases_per_million))+ 
  geom_point(alpha=0.5) +
  geom_smooth(method = "lm") +
  xlab("Population Density (people/ square km)") + ylab("Total cases/million people") + 
  ggtitle("Population density and total cases/million people") +
  theme(plot.title = element_text(hjust = 0.5))
print(pop.den)
```

## Non Pharmaceutical Interventions
##### Overal Government Stringency Score
Overall score is composed of: school closings, workplace closings, cancellation of public events, restrictions on gatherings (by size), public transit closures, stay at home restrictions,internal movement restrictions, international travel controls, public information. 
```{r overall stringency score, echo=TRUE}
npi <- ggplot(reduced, aes(x= npi_stringency_index, y=new_cases_per_million, 
                      color=country))+ 
  geom_boxplot()+
  labs(col="Country")+
  xlab("NPI Stringency Index") + ylab("New Cases/Million People") + 
  ggtitle("New Cases and NPI Stringency Index") +
  theme(plot.title = element_text(hjust = 0.5))
print(npi)
```

### Interventions within Stringency Index
No clear relationship between any specific measures and number of new cases per million. 
```{r NP interventions, echo=TRUE}
#Tried to see the effect of new deaths/million, new cases/million, and daily tpr but didn't see any trend
npi_int <- ggparcoord(reduced, columns = c(22:29, 35, 36, 40),
                  groupColumn = "new_cases_per_million", scale="uniminmax", alphaLines=0.1) +
  xlab("NPI Category") + ylab("Value (scaled between 0 and 1") + 
  ggtitle("Effect of NPIs on New Cases/Million") +
  theme_minimal()+ theme(plot.title = element_text(hjust = 0.5)) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(npi_int)
```

### Mask Policy 
Intensity of mask policy within region.

1 - General recommendation

2 - Mandatory in specific regions of country or specific places (public transport, inside shops)

3 - Mandatory everywhere or universal usage

```{r masks, echo=TRUE}
mask <- ggplot(reduced, aes(x=npi_masks, y=cumulative.tpr))+ 
  geom_boxplot()+
  xlab("Mask Policy") + ylab("Cumulative Test Positivity Rate") + 
  ggtitle("Mask Policy and Cumulative Test Positivity Rate") +
  theme(plot.title = element_text(hjust = 0.5))
print(mask)
```


## Disease Burden
Evaluating the effects of prevalence of diabetes (% of population), smoking (% of group), and cardiovascular deaths (per 1000 people) on the cumulative test positivity rate. 

```{r disease, echo=TRUE}
# Disease burden factors and cTPR
db <- ggparcoord(reduced, columns = 12:15,
                  groupColumn = "cumulative.tpr", scale="uniminmax", alphaLines=0.1) +
  xlab("Disease burden") + ylab("Value") + ggtitle("Disease burden and cumulative TPR") +
  theme_minimal()+ theme(plot.title = element_text(hjust = 0.5)) + theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
print(db)
```

## Mobility Measures

Mobility Trends in % change relative to baseline, from Google Mobility Reports. 

* Retail: restaurants, cafes, shopping centers, theme parks, museums, libraries, and movie theaters.
* Grocery/Pharmacy: grocery markets, food warehouses, farmers markets, specialty food shops, drug stores, and  pharmacies
* Parks: national parks, public beaches, marinas, dog parks, plazas, and public gardens
* Transit: public transport hubs such as subway, bus, and train stations.

```{r mobility, warning=FALSE, echo=TRUE}
pcp <- ggparcoord(reduced, columns = 64:72,
                  groupColumn = "total_cases_per_million", scale="uniminmax", alphaLines=0.1) +
  xlab("mobility factors") + ylab("Value") + 
  ggtitle("Mobility factors and total cases per million people") +
  theme(plot.title = element_text(hjust = 0.5)) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(pcp)
```


## Global Health Security Index
Lower overall GHSindex scores appear to be corellated with lower daily and cumulative tpr. 

```{r ghs index, echo=TRUE}

# Plot 1: Line graph to visualize/compare daily TPR over time
ghs <- ggparcoord(reduced, columns = c(19, 20, 45, 47, 51),
                  groupColumn = "OVERALL_SCORE", scale="uniminmax", alphaLines=0.1) +
  xlab("COVID-19 Data") + ylab("Scaled Values") + ggtitle("GHS Index vs outbreak/testing indicators") +
  labs(col = c("1","2")) +
  theme_minimal()+ theme(plot.title = element_text(hjust = 0.5)) + theme(axis.text.x = element_text(angle = 10, hjust = 0.5)) 
print(ghs)

```

### GHSI Categories
```{r ghs categories, echo=TRUE}

# Plot 1: Line graph to visualize/compare daily TPR over time
ghs <- ggparcoord(reduced, columns = c(97, 154, 191, 234, 273, 312),
                  groupColumn = "cumulative.tpr", scale="uniminmax", alphaLines=0.1) +
  xlab("GHS Indicator") + ylab("Scaled Indicator Value") + ggtitle("Cumulative TPR versus GHS") +
  labs(col = c("1","2")) +
  theme_minimal()+ theme(plot.title = element_text(hjust = 0.5)) 
print(ghs)
```

## Healthcare Access
Assessing relationship between healthcare access indices (using the global health security index) and deaths per million. 

```{r healthcare access, echo=TRUE}
ghs <- ggparcoord(reduced, columns = c(237, 241, 242, 250, 252),
                  groupColumn = "total_deaths_per_million", scale="uniminmax", alphaLines=0.1) +
  xlab("Healthcare Access Indicators") + ylab("Scaled Values") + ggtitle("Deaths per million vs healthcare access measures") +
  labs(col = c("1","2")) +
  theme_minimal()+ theme(plot.title = element_text(hjust = 0.5)) + theme(axis.text.x = element_text(angle = 15, hjust = 0.5)) + scale_x_discrete(labels = function(x) str_wrap(x, width = 5 ))
print(ghs)
```
 
 
# Weather Variables

## Humidity

```{r humidity , echo=TRUE}
# Humidity and total cases (upward trend)
hum <- ggplot(reduced, aes(x=weather_humidity_mean,y=total_cases_per_million))+ 
  geom_point(alpha=0.5) +
  geom_smooth(method = "lm") +
  xlab("Mean humidity (kg water vapour/kg air)") + ylab("Total cases/million people") + 
  ggtitle("Mean humidity and total cases/million people") +
  theme(plot.title = element_text(hjust = 0.5))
print(hum)
```

## Temperature

```{r temperature, echo=TRUE}
temp <- ggplot(reduced, aes(x=weather_temperature_mean,y= new_cases_per_million))+ 
  geom_point(alpha=0.5) +
  geom_smooth(method = "lm") +
  xlab("Mean temperature (°C) ") + ylab("New cases/million people") + 
  ggtitle("Mean temperature and new cases/million people") +
  theme(plot.title = element_text(hjust = 0.5))
print(temp)
```

## Wind

```{r wind, echo=TRUE}
wind <- ggplot(reduced, aes(x=weather_wind_speed_mean, y= new_cases_per_million))+ 
  geom_point(alpha=0.5) +
  geom_smooth(method = "lm") +
  xlab("Mean wind speed (m/s)") + ylab("New cases/million people") + 
  ggtitle("Mean wind speed and new cases/million people") +
  theme(plot.title = element_text(hjust = 0.5))
print(wind)
```

