---
title: "Markdown Practice"
author: "Geetha Jeyapragasan"
date: "July 14 2020"
output:
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(htmltools)
```

```{r load canada, include=FALSE}
canada = read.csv("https://health-infobase.canada.ca/src/data/covidLive/covid19.csv")
canada <- (canada %>%
    mutate (pruid = NULL,prnameFR = NULL) %>% 
    mutate(date = as.Date(date, format= "%d-%m-%Y")) %>%
    mutate (prname = as.factor(prname)) %>%
    mutate (tpr = (numtotal/numtested)*100))
national <- canada %>% filter(prname == "Canada" & !is.na(tpr))
```

## Test Positivity Rates 

The test positivity rate (TPR) is the percentage of tests that are positive. If the positivity rate is too high, it can indicate the region is only testing those who are extremely sick and seeking medical attention, without sampling the wider community. A low test positivity rate can indicate sufficient testing capacity. The WHO has suggested it is safe for regions to reopen if the positivity rate remains ≥ 5% for at least 14 days <https://coronavirus.jhu.edu/testing/testing-positivity>. These measurements focus on real time PCR tests, and do not include serology or antibody tests.  


```{r canada, echo=TRUE}
summary(national)
```

### Canada Test Positivity Rate
```{r tpr, echo=FALSE}
print(
	ggplot(national)
	+ aes(date, tpr)
	+ geom_point()
	+ ylab("Positivity Rate (%)")
)

```


### Canada Testing Rates 

```{r pressure, echo=FALSE}
print(
	ggplot(national)
	+ aes(date, ratetested)
	+ geom_point()
	+ ylab("Test rate (per 1 million people)")
)
```


