---
title: "Test Positivity Rate Writeup"
author: "Geetha Jeyapragasan"
date: "July 14 2020"
output: github_document 
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(htmltools)
```

```{r load canada, include=FALSE}
canada = read.csv("https://health-infobase.canada.ca/src/data/covidLive/covid19.csv")
canada <- (canada %>%
    mutate (pruid = NULL,prnameFR = NULL) %>% 
    mutate(date = as.Date(date, format= "%d-%m-%Y")) %>%
    mutate (prname = as.factor(prname)) %>%
    mutate (tpr = (numtotal/numtested)*100))
national <- canada %>% filter(prname == "Canada" & !is.na(tpr))
```


For COVID-19, the positivity rate  is  used to as a metric to evaluate testing capacity and regime, as well as and severity of an outbreak. If the TPR is too high, it might indicate the region is only testing those who are extremely sick and seeking medical attention, without sampling the wider community. A low TPR can indicate sufficient testing capacity. However due to several factors influencing the test positivity rate, it is not a direct indicator of testing capacity. The test positivity rate has been used as a measure of prevalence for outbreaks that occur in regions with limited or non-existent health surveillance.

### Factors that Affect the TPR 
  1. Type of tests included in metric. Some regions are not transparent in the technologies used for COVID-19 tests, or aggregate PCR and antibody tests making the data less meaningful. 
  2. Whether the numbers are # of performed tests or # of individuals tested. 
  3. Whether negative results are included, and how pending results or inconclusive results are reported. 
  4. Are tests from private labs included? Not all labs report to the central authority, or are specifically not included in the testing data (i.e. US CDC)
  5. Testing eligibility criteria
  6. Whether the time period of cumulative counts if daily counts are not being reported.


### How the TPR is being used and interpreted

* As of May 12, The WHO is currently advising reopening measures based on the TPR. 
The test positivity rate should be below 5% for at least 14 days before reopening, suggesting a
TPR between [3-12%](https://www.who.int/news-room/commentaries/detail/advice-on-the-use-of-point-of-care-immunodiagnostic-tests-for-covid-19) indicates a region is conducting adequate testing. 
* The [CDC opening up plan](https://www.cdc.gov/coronavirus/2019-ncov/downloads/php/CDC-Activities-Initiatives-for-COVID-19-Response.pdf) uses the TPR to assess testing program robustness. The criteria proposes a <20% TPR to enter Phase 1, <15% to enter Phase 2 and <10% to enter Phase 3 (maintaining these values for at least 14 days). 
** Multi-day increases in TPR with stable or increasing testing suggest potential a potential rebound that should be investigated. 
* The TPR was used to compare the temporal trends and mitigation
strategies of COVID-19 between [Seattle and
Washington](https://jamanetwork.com/journals/jama/fullarticle/2766035) .
The paper concluded due to the constant testing volume over the study
period, the TPR trend suggests early and aggressive physical distancing
measures influenced the course of the outbreak. 
* TPR as an good indicator of robust testing and outbreak control was [criticized in
India](https://thewire.in/health/india-covid-19-testing-contradiction-rate)
due to the extremely low testing volume (India is conducting 9.99 tests
per 1000 people while Australia conducts 136 tests per 1000). Though low
testing rates did not fully account for the low TPR, a low TPR in
regions with low testing rates should be interpreted differently than
regions with a high testing rate. 


The WHO recommendation for malaria program managers suggest the use of TPR instead of incidence rates as a measurement if the following 3 factors are inconsistent over time: general outpatient attendance, testing practices, and reporting completeness. Outpatient attendance may be affected by transportation accessibility, user fees, political instability, or general behaviour of treating illnesses at home through over the counter medication or informal drug distributors. If private-for-profit or informal health care providers are taken into account, reporting completeness may be a significant barrier. The TPR is less sensitive to these factors, through it can still be distorted and misinterpreted. 

#### The Effects of creating a TPR Target
*Goodhart's Law: "When a measure becomes a target, it ceases to be a good measure"*

With the focus on test positivity rates as key or even sole indicator of a regions testing and outbreak containment, some regions have developed testing strategies or employed reporting methods to ensure the positivity rate is low. Several states such as Pennsylvania, Texas, Georgia, Vermont, and Virginia have [aggregated their viral and antibody test results](https://www.theatlantic.com/health/archive/2020/05/cdc-and-states-are-misreporting-covid-19-test-data-pennsylvania-georgia-texas/611935/), causing positivity rates across states to appear low. With the state using the low TPR to justify loosening lockdown restrictions, the testing population and types of tests included in the metric should be carefully observed when interpreting the value and used for decision making. California reported a sharp decline in infection rate and positivity rate, but due to [technical issues within the state's data system](https://www.latimes.com/california/story/2020-08-04/californias-coronavirus-test-result-data-may-be-flawed-says-top-health-official) and reporting delays, various counties are severely under-counting community cases. 
Aside from test positivity rates, this has also been noted with other metrics such as exposure time, where the 




### The Global Health Security Index 

The GHS was developed in 2019 to assess the global health capabilities of 195 countries (countries of the International Health Regulations). It is comprised of six categories: prevention,detection and reporting, rapid response, health system, compliance with international norms, and risk environment. Each of these categories are calculated through various indicators, and weighed to generate a score. Looking at the overall score, the U.S is ranked the highest, while their response to COVID-19 has been inadequate in comparison to countries like South Korea, Japan and New Zealand, which were given lower scores. However, when looking into the specific components such as healthcare access, the US ranked quite low (175th out of 195) due to the [lack of guaranteed healthcare access for all citizens](https://www.ghsindex.org/news/the-us-and-covid-19-leading-the-world-by-ghs-index-score-not-by-response/). Identifying which of the indicators/sub-indicators played a crucial role in efficient response, as well as which measures not included in the index were important factors in response. This will allow to better identify how countries can improve their response both for the current pandemic and future global health security risks. 

#### Current Analysis of the GHS Index
* [The Global Health Security Index is not predictive of coronavirus pandemic responses among Organization for Economic Cooperation and Development countries](https://www.medrxiv.org/content/10.1101/2020.07.21.20159061v1). 
** The study (pre-print) found a discrepancy between the GHS index rating
of four indices (i.e., total cases, total deaths, recovery rate, and total tests performed)
and the actual performance of the 36 OECD countries during the pandemic, with an
overestimation of the preparedness of some countries scoring highly on the GHS index
and underestimation of the preparedness of other countries with relatively lower
scores on the GHS index. The U.S and U.K were the highest ranked on the GHS index, but are among the top countries with the highest number of cases per million people. 
* [Rethinking pandemic preparation: Global Health Security Index (GHSI) is predictive of COVID-19 burden, but in the opposite direction](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7207133/). 
** Paper assessed relationship between the index and COVID-19 burden, specifically # of tests, confirmed cases and deaths per million since first confirmed case. The index was plotted against the log transformed values of the outcomes, and a generalized linear model was used to determine the association between the index, confirmed cases and deaths per million. 
** Findings found no association between index and testing rate, a positive association between the index and COVID-19 cases and deaths. This was the opposite of what was expected, though the relationship could have been compounded due to the increased globalization in  most countries with a higher GHS index. This shows the capacity to respond shouldn't be assessed using the overall score, but rather looking into the specific indicators that make a countries response successful. 
* [COVID-19 gives the lie to global health expertise](https://www.thelancet.com/journals/lancet/article/PIIS0140-6736(20)30739-X/fulltext).
** Correspondence in The Lancet discussing the index,specifically analyzing the response of countries that ranked low on the index (comprised of most LMICs).  
** Most African countries were near the bottom of the ranking, but have implemented successful measures in response to COVID-19. Uganda has implemented simplified triage strategies and proactive screening, Rwanda put in handwashing stations at transport hubs, Senegal is using WhatsApp chatbots providing reliable information and rapid testing diagnostics, and Nigeria is using volunteer-staffed call centres and celebrity campaigns to promote responsible actions during the pandemic. Relatively little has been heard about these efforts, though many West and Central African countries made significant improvements to their response following the Ebola outbreak from 2014-2016. 


### Canada Cumulative Test Positivity Rate
```{r tpr, echo=FALSE}
print(
	ggplot(national)
	+ aes(date, tpr)
	+ geom_point()
	+ ylab(" Cumulative Positivity Rate (%)")
)

```


### Canada Testing Rates 

```{r , echo=FALSE}
print(
	ggplot(national)
	+ aes(date, ratetested)
	+ geom_point()
	+ ylab("Cumulative Test rate (per 1 million people)")
)
```

### Cumulative Testing Rate against TPR
```{r, echo=FALSE}
print(
  ggplot(national, aes(ratetested, tpr)) + 
  geom_point() + xlab("Test Rate") + 
  ylab("Cumulative Test Positivity Rate") 
)
```


