---
title: "Test Positivity Rate writeup"
author: "Geetha Jeyapragasan"
date: "July 14 2020"
output: github_document 
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(htmltools)
```

```{r load canada, include=FALSE}
canada = read.csv("https://health-infobase.canada.ca/src/data/covidLive/covid19.csv")
canada <- (canada %>%
    mutate (pruid = NULL,prnameFR = NULL) %>% 
    mutate(date = as.Date(date, format= "%d-%m-%Y")) %>%
    mutate (prname = as.factor(prname)) %>%
    mutate (tpr = (numtotal/numtested)*100))
national <- canada %>% filter(prname == "Canada" & !is.na(tpr))
```

## Test Positivity Rates 

The test positivity rate (TPR) is the percentage of tests that are positive.For COVID-19, this metric is being used to evaluate testing capacity and regime, as well as and severity of an outbreak, (i.e. if two regions had the same number of cases but one region was conducting twice the number of tests, it would suggest the region with the higher positivity rate may be experiencing a more severe outbreak). If the TPR is too high, it might indicate the region is only testing those who are extremely sick and seeking medical attention, without sampling the wider community. A low TPR can indicate sufficient testing capacity. However due to several factors influencing the test positivity rate, it is not a direct indicator of testing capacity. The test positivity rate has been used as a measure of prevalence for outbreaks that occur in regions with limited or non-existent health surveillance. 

### How the TPR is being used and interpreted
AS OF MAY 12, The WHO is currently advising the test positivity rate should be below 5% for at least 14 days before reopening, suggesting a TPR between [3-12%](https://www.who.int/news-room/commentaries/detail/advice-on-the-use-of-point-of-care-immunodiagnostic-tests-for-covid-19) indicates a region is conducting adequate testing. There is a significant difference in TPR between countries, with Australia, South Korea, and Uruguay's positivity rates below 1%, while Argentina and Mexico have a 28% and a 66% positivity rate respectively.

The TPR was used to compare the temporal trends and mitigation strategies of COVID-19 between [Seattle and Washington](https://jamanetwork.com/journals/jama/fullarticle/2766035). The paper concluded due to the constant testing volume over the study period, the TPR trend suggests early and aggressive physical distancing measures influenced the course of the outbreak. TPR as an good indicator of robust testing and outbreak control was [criticized in India](https://thewire.in/health/india-covid-19-testing-contradiction-rate) due to the extremely low testing volume (India is conducting 9.99 tests per 1000 people while Australia conducts 136 tests per 1000). Though low testing rates did not fully account for the low TPR, a low TPR in regions with low testing rates should  be interpreted differently than regions with a high testing rate. 

The WHO recommendation for malaria program managers suggest the use of TPR instead of incidence rates as a measurement if the following 3 factors are inconsistent over time: general outpatient attendance, testing practices, and reporting completeness. Outpatient attendance may be affected by transportation accessibility, user fees, political instability, or general behaviour of treating illnesses at home through over the counter medication or informal drug distributors. If private-for-profit or informal health care providers are taken into account, reporting completeness may be a significant barrier. The TPR is less sensitive to these factors, through it can still be distorted and misinterpreted. 

```{r canada, echo=TRUE}
summary(national)
```

### Canada Test Positivity Rate
```{r tpr, echo=FALSE}
print(
	ggplot(national)
	+ aes(date, tpr)
	+ geom_point()
	+ ylab("Positivity Rate (%)")
)

```


### Canada Testing Rates 

```{r pressure, echo=FALSE}
print(
	ggplot(national)
	+ aes(date, ratetested)
	+ geom_point()
	+ ylab("Test rate (per 1 million people)")
)
```


